### 2015-03-29 swmpr_manu.tex
proofing, finished up to analysis portion

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\vspace*{0.35in}

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{SWMPr: An R package for retrieving, organizing, and analyzing environmental data for estuaries}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Marcus William Beck\textsuperscript{1, *}

@@

The SWMPr package was developed for use with the R statistical programming language and a recent version of R (v3.0.0 or greater) should be installed (see \href{http://cran.r-project.org/}{http://cran.r-project.org/}). The SWMPr package can be installed from \href{https://github.com/fawda123/SWMPr}{GitHub} by executing the following commands at the R terminal.  The package is loaded in the current workspace by using the \texttt{library} command.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{install.packages}\hlstd{(}\hlstr{'devtools'}\hlstd{)}
\hlkwd{library}\hlstd{(devtools)}
\hlkwd{install_github}\hlstd{(}\hlstr{'fawda123/SWMPr'}\hlstd{)}
\hlkwd{library}\hlstd{(SWMPr)}
\end{alltt}
\end{kframe}
\end{knitrout}

The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: {\bf{\it retrieving}}, {\bf{\it organizing}}, and {\bf{\it analyzing}}.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.  Functions for organizing and analyzing the data provide methods for working with the \texttt{swmpr} object class.  An additional class of functions, termed `miscellaneous', are also included as helpers for the main functions.  The following describes a general approach for using each category of functions based on a standard data workflow.

@@

Three approaches can be used for retrieving \ac{SWMP} data for import into R.  First, functions from the SWMPr package can be used to import the data directly from the online server using \ac{CDMO} web services (\Cref{tab:retrieve}).  To do so, the IP address for the computer making the request must be registered with \ac{CDMO}.  This can be done by following instructions \href{http://cdmo.baruch.sc.edu/webservices.cfm}{here}.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.

% table for retrieval functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
\begin{table}[!tbp]
\caption{Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?all\_params} at the command line).\label{tab:retrieve}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{all\_params}&Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.\tabularnewline
\texttt{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter.  Maximum of 1000 records. Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.\tabularnewline
\texttt{import\_local}&Import files from a local path.  The files must be in a specific format, specifically those returned from the \ac{CDMO} using the zip downloads option for a reserve.\tabularnewline
\texttt{import\_remote}&Import site data from an independent server. These files have been downloaded from \ac{CDMO} up to December 2014, processed using functions in this package, and uploaded to an Amazon server for quicker import into R.\tabularnewline
\texttt{single\_param}&Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.\tabularnewline
\texttt{site\_codes}&Metadata for all stations, wrapper to \texttt{exportStationCodesXMLNew} function on web services.\tabularnewline
\texttt{site\_codes\_ind}&Metadata for all stations at a single site, wrapper  to \texttt{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# retrieve metadata for all sites}
\hlkwd{site_codes}\hlstd{()}

\hlcom{# retrieve metadata for a single site}
\hlkwd{site_codes_ind}\hlstd{(}\hlstr{'apa'}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

Due to rate limitations on the \ac{CDMO} server, the retrieval functions return a limited number of records.  The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} loops) to obtain longer time series.  Data retrieval functions to access the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}.  These are functions that call the existing web protocol methods on the CDMO web services.  \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# all parameters for a station, most recent}
\hlkwd{all_params}\hlstd{(}\hlstr{'hudscwq'}\hlstd{)}

\hlcom{# get all parameters within a date range}
\hlkwd{all_params_dtrng}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlkwd{c}\hlstd{(}\hlstr{'09/10/2012'}\hlstd{,} \hlstr{'02/8/2013'}\hlstd{))}

\hlcom{# get single parameter within a date range}
\hlkwd{all_params_dtrng}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlkwd{c}\hlstd{(}\hlstr{'09/10/2012'}\hlstd{,} \hlstr{'02/8/2013'}\hlstd{),}
  \hlkwc{param} \hlstd{=} \hlstr{'do_mgl'}\hlstd{)}

\hlcom{# single parameter for a station, most recent}
\hlkwd{single_param}\hlstd{(}\hlstr{'hudscwq'}\hlstd{,} \hlstr{'do_mgl'}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

A second approach for importing \ac{SWMP} data into R is using the \texttt{import\_remote} function.  This method imports site data into R from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in RData binary format.  This approach is advantageous because the files can be imported very quickly and the computer's IP address does not need to be registered.  The files are hosted via third party web services and include almost all available \ac{SWMP} data for each site up to December 2014.  Always use the \ac{CDMO} for the most recent data.  The files can be accessed by copying the URL to a web browser with the site name and file extension (e.g., \href{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}, see the \nameref{supp_info}) or by using the function as follows.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## import a file from an independent server}
\hlkwd{import_remote}\hlstd{(}\hlstr{'acebbwq'}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

A third approach is to use the \texttt{import\_local} function to import data into R that are locally available after downloading outside of R (\href{http://cdmo.baruch.sc.edu/get/landing.cfm}{see here}).  This approach may be best for larger, more customized data requests to \ac{CDMO}. Data can be retrieved from the \ac{CDMO} several ways.  The \texttt{import\_local} function is designed for data from the \href{http://cdmo.baruch.sc.edu/aqs/zips.cfm}{zip downloads} feature in the advanced query section of the \ac{CDMO}. The function may also work using data from the \href{http://cdmo.baruch.sc.edu/get/export.cfm}{data export system}, but this feature has not been extensively tested.  The zip downloads feature is an easy way to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \texttt{import\_local} function can be used after the folder is decompressed.

Occasionally, duplicate time stamps are present in the raw data.  The \texttt{import\_local} function handles duplicate entries differently depending on the data type (water quality,  weather, or nutrients).  For water quality and nutrient data, duplicate time stamps are simply removed.  Note that nutrient data often contain replicate samples with similar but not duplicated time stamps within a few minutes of each other.  Replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.  Weather data prior to 2007 may contain duplicate time stamps at frequencies for 60 (hourly) and 144 (daily) averages, in addition to 15 minute frequencies.  Duplicate values that correspond to the smallest value in the frequency column (15 minutes) are retained.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# import data for apaebmet that you downloaded}

\hlcom{# this is an example path with the csv files, change as needed}
\hlstd{path} \hlkwb{<-} \hlstr{'C:/my_path/'}

\hlcom{# import, do not include file extension}
\hlkwd{import_local}\hlstd{(path,} \hlstr{'apaebmet'}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

All data retrieval functions return a \texttt{swmpr} object that includes relevant data and several attributes describing the dataset.  The data include a \texttt{datetimestamp} column in the appropriate timezone for a station.  Note that the \texttt{datetimestamp} is standard time for each timezone and does not include daylight savings. Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg \ac{QAQC} columns if returned from the initial data request.  The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}) \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).  Attributes of a \texttt{swmpr} object can be viewed as follows.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# import binary data}
\hlkwd{data}\hlstd{(apadbwq)}
\hlstd{dat} \hlkwb{<-} \hlstd{apadbwq}

\hlcom{# verify that dat is swmpr class}
\hlkwd{class}\hlstd{(dat)}
\end{alltt}
\begin{verbatim}
## [1] "swmpr"      "data.frame"
\end{verbatim}
\begin{alltt}
\hlcom{# all attributes of dat}
\hlkwd{names}\hlstd{(}\hlkwd{attributes}\hlstd{(dat))}
\end{alltt}
\begin{verbatim}
## [1] "names"       "row.names"   "class"       "station"    
## [5] "parameters"  "qaqc_cols"   "date_rng"    "timezone"   
## [9] "stamp_class"
\end{verbatim}
\begin{alltt}
\hlcom{# a single attribute of dat}
\hlkwd{attr}\hlstd{(dat,} \hlstr{'station'}\hlstd{)}
\end{alltt}
\begin{verbatim}
## [1] "apadbwq"
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# available methods for swmpr class}
\hlkwd{methods}\hlstd{(}\hlkwc{class} \hlstd{=} \hlstr{'swmpr'}\hlstd{)}
\end{alltt}
\begin{verbatim}
##  [1] aggregate.swmpr       aggregate_metab.swmpr comb.swmpr           
##  [4] decomp.swmpr          decomp_cj.swmpr       ecometab.swmpr       
##  [7] hist.swmpr            lines.swmpr           na.approx.swmpr      
## [10] plot.swmpr            plot_metab.swmpr      plot_summary.swmpr   
## [13] qaqc.swmpr            qaqcchk.swmpr         rem_reps.swmpr       
## [16] setstep.swmpr         smoother.swmpr        subset.swmpr
\end{verbatim}
\end{kframe}
\end{knitrout}

Example data as raw, comma-separated files have not been included in the package due to size limitations.  However, a sample dataset can be \href{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}{downloaded} for use with the examples below.  This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.  Processed verions of the raw data are included with the package as binary data files (RData) to decrease processing times with the examples.  Information for each binary file can be viewed as follows.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# view help files for complementary data}
\hlcom{# all files are samples from Apalachicola Bay}

\hlcom{# cat point station, nutrients}
\hlopt{?}\hlstd{apacpnut}

\hlcom{# cat point station, water quality}
\hlopt{?}\hlstd{apacpwq}

\hlcom{# dry bar station, water quality}
\hlopt{?}\hlstd{apadbwq}

\hlcom{# east bay station, weater}
\hlopt{?}\hlstd{apaebmet}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions. The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types (\Cref{tab:organize}).

The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags (see the \ac{CDMO} descriptions \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}).  Each parameter in the \texttt{swmpr} data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.  Values in the QAQC column specify a flag from -5 to 5.  Generally, only data with the  \texttt{0} \ac{QAQC} flag should be used, which is the default option for the \texttt{QAQC} function.  Data that do not satisfy \ac{QAQC} criteria are converted to \texttt{NA} values.   Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.

% table for organize functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?comb.swmpr} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE)%
\begin{table}[!tbp]
\caption{Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?comb.swmpr} at the command line).\label{tab:organize}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{comb.swmpr}&Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.\tabularnewline
\texttt{qaqc.swmpr}&Remove \ac{QAQC} columns and remove data based on \ac{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \ac{QAQC} columns are present. \tabularnewline
\texttt{qaqcchk.swmpr}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different \ac{QAQC} flags used by \ac{CDMO}.  The output is used to inform further processing but is not used explicitly.\tabularnewline
\texttt{rem\_reps.swmpr}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
\texttt{setstep.swmpr}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.  The function is used in \texttt{comb.swmpr} and can also be used with individual stations.\tabularnewline
\texttt{subset.swmpr}&Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# qaqc screen for a swmpr object, retain only '0'}
\hlkwd{qaqc}\hlstd{(dat)}

\hlcom{# retain all data regardless of flag}
\hlkwd{qaqc}\hlstd{(dat,} \hlkwc{qaqc_keep} \hlstd{=} \hlkwa{NULL}\hlstd{)}

\hlcom{# retain only '0' and '-1' flags}
\hlkwd{qaqc}\hlstd{(dat,} \hlkwc{qaqc_keep} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{,} \hlopt{-}\hlnum{1}\hlstd{))}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# view the number of observations in each QAQC flag}
\hlkwd{qaqcchk}\hlstd{(dat)}
\end{alltt}
\end{kframe}
\end{knitrout}

Raw nutrient data obtained from the \ac{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call including treatment of NA values should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# get nutrient data}
\hlkwd{data}\hlstd{(apacpnut)}
\hlstd{swmp1} \hlkwb{<-} \hlstd{apacpnut}
\hlstd{swmp1} \hlkwb{<-} \hlkwd{qaqc}\hlstd{(swmp1)}

>>>>

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\vspace*{0.35in}
\urlstyle{same}

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{SWMPr: An R package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Marcus William Beck\textsuperscript{1, *}

@@

The SWMPr package was developed for use with the R ($\geq$v3.0.0) statistical programming language \cite{RDCT14}. The SWMPr package can be installed from GitHub\footnote{\url{https://github.com/fawda123/SWMPr}} by executing the following commands at the R console prompt.  The package is loaded in the workspace using the \texttt{library} command.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> install.packages('devtools')
> library(devtools)
> install_github('fawda123/SWMPr')
> library(SWMPr)
\end{verbatim}
\end{kframe}
\end{knitrout}

The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class following standard S3 documentation methods \cite{Wickham14}.  Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes a general approach for using each category of functions.

@@

Three approaches can be used to import \ac{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using \ac{CDMO} web services. To do so, the IP address for the computer making the request must be registered with \ac{CDMO}.\footnote{Follow instructions here: \url{http://cdmo.baruch.sc.edu/webservices.cfm}}  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.

% table for retrieval functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE)%
\begin{table}[!tbp]
\caption{Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?all\_params} at the command line).\label{tab:retrieve}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{all\_params}&Retrieve up to 100 records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.\tabularnewline
\texttt{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter.  Maximum of 1000 records. Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.\tabularnewline
\texttt{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the \ac{CDMO} using the zip downloads option.\tabularnewline
\texttt{import\_remote}&Import site data from an independent server. These files have been downloaded from \ac{CDMO} up to December 2014, processed using SWMPr functions, and uploaded to an Amazon server for quicker import into R.\tabularnewline
\texttt{single\_param}&Retrieve up to 100 records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.\tabularnewline
\texttt{site\_codes}&Metadata for all stations, wrapper to \texttt{exportStationCodesXMLNew} function on web services.\tabularnewline
\texttt{site\_codes\_ind}&Metadata for all stations at a single site, wrapper  to \texttt{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # retrieve metadata for all sites
> site_codes()
> 
> # retrieve metadata for a single site
> site_codes_ind('apa')
\end{verbatim}
\end{kframe}
\end{knitrout}

Due to rate limitations on the \ac{CDMO} server, the retrieval functions return a limited number of records.  The functions are more useful for evaluating short time periods, although these functions could be used iteratively (i.e., with \texttt{for} loops) to obtain longer time series.  Data retrieval functions to access the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # all parameters for a station, most recent
> all_params('hudscwq')
> 
> # get all parameters within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'))
> 
> # get single parameter within a date range
> all_params_dtrng('hudscwq', c('09/10/2012', '02/8/2013'), 
>   param = 'do_mgl')
> 
> # single parameter for a station, most recent
> single_param('hudscwq', 'do_mgl')
\end{verbatim}
\end{kframe}
\end{knitrout}

A second approach for importing \ac{SWMP} data into R uses the \texttt{import\_remote} function.  This method imports site data from an independent server where \ac{SWMP} data have been downloaded previously and uploaded as \texttt{swmpr} objects in RData binary format.  This approach is advantageous because the files can be imported very quickly and an IP address does not need to be registered.  The files include almost all available \ac{SWMP} data for each site up to December 2014.  However, the files will not be regularly updated and \ac{CDMO} should always be used for the most recent and accurate data.  The files can be accessed by copying the URL to a web browser with the site name and file extension (e.g., \href{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}{https://s3.amazonaws.com/swmpalldata/acebbnut.RData}, see the \nameref{supp_info}) or by using the function as follows.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import a file from an independent server
> import_remote('acebbwq')
\end{verbatim}
\end{kframe}
\end{knitrout}

A third approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.  This approach is most appropriate for large, specific data requests. The \texttt{import\_local} function is designed for data from the zip downloads\footnote{\url{http://cdmo.baruch.sc.edu/aqs/zips.cfm}} feature in the advanced query section of the \ac{CDMO}. The function may also work using data from the data export system,\footnote{\url{http://cdmo.baruch.sc.edu/get/export.cfm}} although this feature has not been extensively tested.  The zip downloads feature is an easy way to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \texttt{import\_local} function can be used after the folder is decompressed.

Occasionally, duplicate time stamps are present in the raw data.  The \texttt{import\_local} function handles duplicate entries differently depending on the data type (water quality,  weather, or nutrients).  For water quality and nutrient data, duplicate time stamps are simply removed.  Note that nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.  Replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.  Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.  Only duplicate values at 15 minutes are averaged for weather data.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import local data for apaebmet
> 
> # this is an example path with the decompressed csv files
> path <- 'C:/my_path/'
> 
> # import, do not include file extension
> import_local(path, 'apaebmet') 
\end{verbatim}
\end{kframe}
\end{knitrout}

All data retrieval functions return a \texttt{swmpr} object that includes relevant data and several attributes describing the dataset.  The data include a \texttt{datetimestamp} column in the appropriate timezone for a station.  Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg \ac{QAQC} columns if returned from the initial data request.  The attributes for a \texttt{swmpr} object include \texttt{names} of the dataset, \texttt{row.names} of the dataset, \texttt{class} (character string indicating \texttt{swmpr} and \texttt{data.frame}), \texttt{station} (7 or 8 characters identifying the station),  \texttt{parameters} (character vector of data columns, e.g., \texttt{`do\_mgl'}), \texttt{qaqc\_cols} (logical \texttt{T} or \texttt{F} if present or not), \texttt{date\_rng} (\texttt{POSIXct} vector of minimum/maximum dates), \texttt{timezone} (text string in country/city format, standard time only), and \texttt{stamp\_class} (class of \texttt{datetimestamp} vector, \texttt{POSIXct} or \texttt{Date}).  Attributes of a \texttt{swmpr} object can be viewed as follows.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # import data
> dat <- import_remote('apadbwq')
>  
> # view all attributes of dat
> attributes(dat)
>  
> # view a single attribute of dat
> attr(dat, 'station')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # view available methods for swmpr class
> methods(class = 'swmpr')
\end{verbatim}
\end{kframe}
\end{knitrout}

A sample dataset can be downloaded for use with the examples below.\footnote{\url{https://s3.amazonaws.com/swmpexdata/zip_ex.zip}}  This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.  Processed verions of the raw data are also included with the package as binary data files (RData).  These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).

@@

The organize functions are used to clean or prepare the imported data for analysis, including viewing and removal of QAQC flags, subsetting, combining replicate nutrient observations, creating a standardized time series, and combining data of different types (\Cref{tab:organize}).

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags.\footnote{\ac{CDMO} flags: \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}} Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the \ac{QAQC} flag that was assigned by \ac{CDMO} during initial processing.  The \ac{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \ac{QAQC} flag (i.e., passed initial \ac{CDMO} checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

% table for organize functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?comb.swmpr} at the command line).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE)%
\begin{table}[!tbp]
\caption{Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?comb.swmpr} at the command line).\label{tab:organize}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{comb.swmpr}&Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.\tabularnewline
\texttt{qaqc.swmpr}&Remove \ac{QAQC} columns and remove data based on \ac{QAQC} flag values for a \texttt{swmpr} object.  Only applies if \ac{QAQC} columns are present. \tabularnewline
\texttt{qaqcchk.swmpr}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to different \ac{QAQC} flags used by \ac{CDMO}.  The output can be used to inform further processing.\tabularnewline
\texttt{rem\_reps.swmpr}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
\texttt{setstep.swmpr}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.  The function is used in \texttt{comb.swmpr} and can also be used with individual stations.\tabularnewline
\texttt{subset.swmpr}&Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.\tabularnewline
\hline
\end{tabular}\end{center}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # qaqc screen for a swmpr object, retain only '0'
> qaqc(dat)
> 
> # retain all data regardless of flag
> qaqc(dat, qaqc_keep = NULL)
> 
> # retain only '0' and '-1' flags
> qaqc(dat, qaqc_keep = c(0, -1))
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # view the number of observations in each QAQC flag
> qaqcchk(dat)
\end{verbatim}
\end{kframe}
\end{knitrout}

Raw nutrient data obtained from the \ac{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call, including treatment of \texttt{NA}, values should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \ac{QAQC} columns are present.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
> # get nutrient data
> data(apacpnut)
> dat <- apacpnut
> dat <- qaqc(dat)
> 
> # remove replicate nutrient data
> rem_reps(dat)
> 
> # use different function to aggregate replicates
> func <- function(x) max(x, na.rm = T)
> rem_reps(dat, FUN = func)
\end{verbatim}
\end{kframe}
\end{knitrout}
